Neutral models of \textit{de novo} gene emergence suggest that gene evolution has a preferred trajectory

### 2023-02-16 Manuscripts/Response_to_Reviewers_Ravi2023MBE.tex

\textbf{1) Basal transcription:} The authors repeatedly mention that in order for transcription to take place there is a need for a promoter to evolve. Much empirical work in the field of de novo gene emergence has shown how basal levels of transcription or expression due to chromatin rearrangements might prime the functionalization of de novo genes

\begin{itemize}\setlength{\itemsep}{-5pt}
\item https://genome.cshlp.org/content/20/10/1313
\item https://elifesciences.org/articles/47138
\item https://academic.oup.com/mbe/article/37/4/1165/5679780
\item https://genome.cshlp.org/content/28/11/1675.short
\item https://elifesciences.org/articles/09977
\end{itemize}

Because the presence/absence of transcription is a fundamental aspect of the model in this paper, I think this paradigm needs to be integrated into it for this study to be relevant given the knowledge we now have.
}

We thank the reviewer for this suggestion and for pointing us to relevant studies. It is indeed true that low level of transcription exists even in the absence of a core promoter motif. Furthermore enhancers can also initiate transcription in their proximal regions. However, we also note that vast majority of RNAs arising due to pervasive transcription (including the enhancer associated eRNAs), are degraded in the nucleus and thus do not reach the cytoplasm where they can be translated. Therefore, we think that a polyadenylation signal would still be a necessary feature for determining transcript length and stability. We further rationalize this argument by two observations. First, polyA signals are 10-fold less enriched in coding regions (0.0004) than in intergenic regions (0.004). This suggests that polyA signals within the gene body (including ORF) can be disruptive. Second, many studies on \textit{de novo} gene discovery to our knowledge, employ a RNAseq library preparation kit that amplifies polyadenylated RNAs (\href{https://doi.org/10.7554/eLife.09977}{Neme and Tautz, 2016}; \href{https://doi.org/10.7554/eLife.47138}{Witt \textit{et al.}, 2019}; \href{https://doi.org/10.1093/molbev/msz300}{Majic and Payne, 2019}; \href{https://doi.org/10.1038/s41467-021-20911-3}{Blevins \textit{et al.}, 2021}). Thus most of the discovered \textit{de novo} genes indeed have a polyA tail (and must also have a polyadenylation signal). We now mention both these observations in the main text (lines XXX).

We next focus on sequences that initiate transcription. Although several DNA sequence motifs corresponding to enhancers and core promoters have been identified, there is no set of sequence motifs that is deemed to be absolutely necessary. The best understood, and the most common core promoter motifs are TATA-box and Inr, which we included in our original analysis. We further note that although some sequences are indeed associated with enhancers, the actual functional enhancers are largely determined by epigenetic marks that vary significantly in different cell types, developmental stages, and during different environmental conditions. One common feature of both promoters and enhancers is that they both are characterized by a nucleosome depleted region (\href{https://doi.org/10.1038/s41576-019-0173-8}{Andersson and Sandelin, 2020}). Overall, we surmise that transcription can be initiated at any nucleosome depleted region but the stability of such transcripts is determined by polyA signals.

We assume that such nucleosome free regions are uniformly distributed in open chromatin. With the condition that a \textit{de novo} gene emerges in open chromatin regions, we now calculated the probability of transcript emergence based on poly-A signals alone, while ignoring the requirement of a promoter. We first asked if such an assumption would significantly change the results presented in our first draft (Figures 1 -- 3). We found that these results do not change significantly. The fundamental inferences remain the same. If we had data on every promoter/enhancer sequence, and if we incorporate them in the model, then the exact RNA probability values should lie in between those estimated by the promoterless model and TATA/Inr-only model. In the revised manuscript we now present the results from the promoterless model, and provide a suitable justification in the text (lines XXX). We include the old results based on TATA/Inr-only model in the supplementary data (Fig SXX).

{\itshape \textbf{2) Non-neutral background:} de novo ORFs might emerge in non-coding regions that still encode for important cellular functions, such as introns or enhancers. This would mean that some mutations that lead to the functionalization of a de novo peptide might be unviable because they temper with the functions of the genomic regions where they arise. This might make some sequences harder or easier to evolve than others because some specific genomic positions will be locked in to a given nucleotide or arrangement of nucleotide. The way this could be integrated into the model is by considering what is the percentage of sites that tends to be strongly constrained or conserved between different species.}

The reviewer raised an important concern. To address this concern, we performed an analysis of total number of available sites and the probability of gene emergence, focusing on the model organism \textit{D.melanogaster}. To this end, we first downloaded the genomic coordinates of all \textit{cis}-regulatory motifs from \href{http://redfly.ccr.buffalo.edu/}{REDfly database}. Next, we calculated the lengths of intergenic sequences that are not interrupted by any \textit{cis}-regulatory motif, and estimated the total number of sites available for genes of different lengths to emerge. Finally, we calculated the total gene gain probability by multiplying the total number of feasible sites and per-site gene gain probability. We now briefly describe these findings in the main text, and in greater detail in the supplementary data (Figure SXX).

We would like to note that in the original analysis, we calculate the probabilities of gene loss and gain, at one genomic locus. We also assume that this locus exists in an intergenic region devoid of any functional sequences. Therefore the original results will not be dependent on the number of feasible loci (or specific genomes). We have now clarified this assumption in the revised manuscript (lines XXX).

@@

{\itshape \textbf{3) Structured probabilities:} The authors rightly used empirical estimations of CG content to study the probability of functionalizing or defunctionalizing mutations. But It's not just CG content, it might be important to consider the frequency or presence/absence of genomic patterns like CpG islands or microsatellites. The existence of these features will strongly influence the likelihood of getting specific kinds of mutations, which will be important for the biases in evolution as well as the propensities in protein evolution that the authors tried to study with their model. Not to mention that these features will have different probabilities of evolving transcription on top of the coding sequence.}

We thank the reviewer for this valuable suggestion. We agree with them that nucleotide distribution is not uniform and therefore the probabilities of different DNA sequences cannot be estimated accurately from GC content alone. We also note that sequence composition can be highly variable between different species, as well as between different genomic regions in the same species. Thus the most accurate way to estimate the rate of gene evolutionary dynamics would be to analyse each genomic locus individually. Although such an analysis would be accurate, it would be difficult to generalize. To keep a balance between accuracy and generalizability of our results, we repeated our original calculations of various probability using nucleotide distribution estimates from intergenic regions of \textit{Drosophila melanogaster}  (flybase release 6.49). Specifically, we estimated the frequencies of all hexamers in the intergenic sequences using a sliding window. We ignored the hexamers that contained an ambiguous nucleotide character (non ATGC). From the frequency distribution of all the 4096 hexamers, we estimated the probability of finding (also gaining and losing) the polyadenylation signal sequences (that are 6 nucleotides long). Similarly, we calculated the probabilities of start and stop codons, using a distribution of trimers estimated from the \textit{D.melanogaster} intergenic regions. With these probabilities, we repeated the analyses shown in the original figures 1A, 2A, 2C and 3A.

We found that the the results from actual nucleotide distributions indeed differed from the ones obtained using total GC content of \textit{D.melanogaster} intergenic genome (0.403). We now include these results in the main manuscript along with the old figures. We also explain the details of the re-analysis and its rationale (as described in this response), in the main text. (lines XXX)

{\itshape \textbf{4) Population genetics:} Ultimately a de novo gene needs to increase in frequency in a population if it is going to be maintained in the genome. The probability of fixation of a neutral allele is 1/2N, but how is the birth/death mutational dynamics of de novo genes that the authors are exploring affected by the demography of species? I believe that drift might be the primary source of loss of de novo neutral genes, much more than mutational defunctionalization. Or do the authors want to disregard this and just focus on the macroevolutionary retention of de novo genes in phylogenies? If that is the case, this could be discussed, maybe not necessarily incorporated into the model.}

### 2023-02-17 Manuscripts/Response_to_Reviewers_Ravi2023MBE.tex

Because of its theoretical approach this paper is a good addition to the de novo gene literature, however, I think the paper will be more relevant if it would either:

\begin{enumerate}
\item offer a more exhaustive contextualization of the model and the field, emphasizing why the model is showing results that cannot be addressed with empirical data, or

\item it would evaluate some of their results with empirical genetic diversity or experimental evolution datasets (such as https://academic.oup.com/mbe/article/39/2/msab368/6491261) to test whether the rates of ORF gain or loss within a population match their predictions. As it stands, although formalizing these probabilities theoretically has the potential to be an important contribution, given its simplifications, I believe the model is not too informative of the realistic potential of its predictions.
\end{enumerate}
}

anna's data (table 4). Can be estimated by total rnagain * total sites * generations per year * \% open chromatin (1-6)* \% open chromatin in intergenic region (35)

@@

\textbf{1) Basal transcription:} The authors repeatedly mention that in order for transcription to take place there is a need for a promoter to evolve. Much empirical work in the field of de novo gene emergence has shown how basal levels of transcription or expression due to chromatin rearrangements might prime the functionalization of de novo genes

\begin{itemize}\setlength{\itemsep}{-5pt}
\item https://genome.cshlp.org/content/20/10/1313
\item https://elifesciences.org/articles/47138
\item https://academic.oup.com/mbe/article/37/4/1165/5679780
\item https://genome.cshlp.org/content/28/11/1675.short
\item https://elifesciences.org/articles/09977
\end{itemize}

Because the presence/absence of transcription is a fundamental aspect of the model in this paper, I think this paradigm needs to be integrated into it for this study to be relevant given the knowledge we now have.
}

We thank the reviewer for this suggestion and for pointing us to relevant studies. It is indeed true that low level of transcription exists even in the absence of a core promoter motif. Furthermore enhancers can also initiate transcription in their proximal regions. However, we also note that vast majority of RNAs arising due to pervasive transcription (including the enhancer associated eRNAs), are degraded in the nucleus and thus do not reach the cytoplasm where they can be translated. Therefore, we think that a polyadenylation signal would still be a necessary feature for determining transcript length and stability. We further rationalize this argument by two observations. First, polyA signals are 10-fold less enriched in coding regions (0.0004) than in intergenic regions (0.004). PolyA signals are also 3 times less enriched in \textit{D.melanogaster} RNAs than in the intergenic regions. Furthermore, the polyA signals in RNAs are usually located at the end of the RNA. These observations suggests that polyA signals within the gene body (including ORF) can be disruptive. Second, many studies on \textit{de novo} gene discovery to our knowledge, employ a RNAseq library preparation kit that amplifies polyadenylated RNAs (\href{https://doi.org/10.7554/eLife.09977}{Neme and Tautz, 2016}; \href{https://doi.org/10.7554/eLife.47138}{Witt \textit{et al.}, 2019}; \href{https://doi.org/10.1093/molbev/msz300}{Majic and Payne, 2019}; \href{https://doi.org/10.1038/s41467-021-20911-3}{Blevins \textit{et al.}, 2021}). Thus most of the discovered \textit{de novo} genes indeed have a polyA tail (and must also have a polyadenylation signal). We now mention both these observations in the main text (lines XXX).
